Tracing prices: A flow-based cost allocation for optimized power systems

### 2020-10-22 paper/flow_based.tex
follow up

\newcommand{\allocatecost}[1][n \rightarrow i]{\cost_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s,t]{\capexgeneration_{#1}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell,t]{\capexflow_{#1}}
\newcommand{\allocatecapexstorage}[1][n \rightarrow r, t]{\capexstorage_{#1}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\opex_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\emissioncost_{#1,t}}
\newcommand{\allocatescarcitycost}[1][n \rightarrow i]{\scarcitycost_{#1,t}}
\newcommand{\allocatesubsidycost}[1][n \rightarrow i]{\subsidycost_{#1,t}}

@@

    \begin{abstract}
        Cost minimization of power systems leads to perfectly scheduled operations and investments while securing all power demands. 
        System needs reports from regional, national and international agencies increasingly rely on cost minimization techniques. However, cost impact to single regions and thefore the question of who profits to what extent from an investment is often neglected. The literature provides multiple cost allocation methods dedicated to this problem, yet they hardly suit for large networks and an integration of all costs. 
        This work introduces a peer-to-peer allocation of all system costs in optimized models. Combining basic relations from economic optimization with flow allocation, its cost assignments locally constraint while still being aligned with the Locational Marginal Prices at the optimum. It is showcased by means of an optimized near-future scenario of the German power system. 
        \vtop{%
         \vskip0pt
          \hbox{%
          \includegraphics[width=\linewidth]{graphical_abstract.png}
          }%
        }

@@

\subsection*{Highlights}
\begin{itemize}
    \item In long-term equilibria a network asset gains back its variable and investment cost from its operation based revenue
    \item Flow tracing is used as a basis to allocate the operation of assets to consumers. 
    \item Allocated flows have to be reshaped such that the Kirchhoff Current Law and the Kirchhoff Voltage Law are respected. 
    \item Using operational and shadow prices from constraints, all costs are assigned in an P2P manner.
    \item The cost assignments are locally constraint and in alignment with the nodal pricing scheme.    
    % \item Upper and lower capacity expansion limits modify the revenue per asset, which however correctly reflects in the cost allocation
    \item In a low-carbon scenario for Germany regions with high renewable potentials profit from investment compensated by remote buses.   
\end{itemize}

@@

Today's power systems are subject to a deep and ongoing transformation. The needed shift to from controllable to weather-driven power generation, the constant improvement and innovation of technology require rigorous system planning and international cooperation \cite{pfenninger_energy_2014,schlachtberger_benefits_2017}. The core of the problem manifests in the costs. Firstly, these should be as small as possible while meeting ecological and sociological standards. Secondly, they must be distributed in a fair and transparent manner. Cost-driving aspects or market players should be detected and addressed appropriately. 
In this context, power system modelling has increasingly gained attention during the last years \cite{pfenninger_energy_2014,bazmi_sustainable_2011,pereira_generation_2017}. Many studies show how cost-efficiency and renewable energy are brought together. However, the question of how and, more importantly, upon what grounds costs are distributed was often left open. \\
This study bridges this gap. On the basis of an optimized network the operational state of each time step and corresponding characteristics are in detailed considered to allocate all system costs. The presented approach builds upon two basic concepts: Firstly, the revenue of an network asset at the cost-optimum exactly matches its operational and capital expenditures, also known as the zero-profit condition. Secondly, flow tracing, following Bialek's Average Participation (AP) \cite{bialek_tracing_1996}, can be used to assigns the operation of single network assets to consumers in an plausible and locally constraint manner. 
Combining these two relations allows for an transparent allocation of all operational (OPEX) and capital expenditures (CAPEX) to the consumers in the network. \\
Flow-based cost allocation was already discussed and performed in various works \cite{galiana_transmission_2003,shahidehpour_market_2002,meng_investigation_2007,tranberg_flow-based_2018,nikoukar_transmission_2012,arabali_pricing_2012,wu_locational_2005}. Shahidehpour et al. give a deep insight into allocating congestion cost and transmission investments to market participants \cite{shahidehpour_market_2002} with different allocation techniques. Based on Generation Shift Factors, \ie the marginal contribution of generators to a flow on line, the Locational Marginal Price (LMP) can be represented as a superposition of LMP at the reference bus, the price for congestion and a price for losses. The approach in \cite{meng_investigation_2007} expands this relation for contributions based on the AP scheme, which allows for an accurate estimation of the LMP, however does not reflect the exact LMP of an optimized network. A similar approach was taken in \cite{tranberg_flow-based_2018} where electricity prices of a non-optimal power dispatch were allocated using flow tracing.

@@

The first section which presents the full method splits into the presentation of: the zero-profit condition for different kind of assets (\cref{sec:zero_profit_rule});  the AP allocation scheme and resulting allocation from assets to consumers (\cref{sec:dispatch_allocations}); the cost allocation and the impact of additional constraints (\cref{sec:general_scheme,sec:design_constraints}); a numerical example (\cref{sec:numerical_example}). The second part in \cref{sec:application_case} applies the cost allocation to a optimized German power system with high shares of renewable power sources and evaluates the allocated costs.

@@

\subsubsection*{Generators}
\begin{subequations}
Let $S \subseteq I$ be the set of generators in the network, such that $\generation = s_{s,t}$ describes the power production of generator $s \in S$. The OPEX occasioned by generator $s$ is given by a cost-weighted sum of the production, thus 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
In case a fix price for emissions $\emissionprice$ in \euro\, per tonne-CO$_2$ equivalents, is assumed, a further the cost term per generator $s$     
\begin{align}
 \emissioncost_s = \emissionprice \, \sum_t  \emission \, \generation
\end{align}
adds to $\totalcost$. Here, $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour\, of generator $s$. In contrast to OPEX and emission costs, the CAPEX of $S$ are not a function of the production $\generation$, but of the actual installed capacity $\capacitygeneration$ of generator $s$. In the optimization it limits the generation $\generation$ in the form of 
\begin{align}
\generation - \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}
\end{align}
The constraint yields a shadow-price of $\muuppergeneration$ in literature often denoted as the Quality of Supply \cite{schweppe_spot_1988}. It can be interpreted as the price per MW that \cref{eq:upper_generation_capacity_constraint} imposes to the system. If $\muuppergeneration$  is bigger than zero, the constraint is binding, which pushes investments in $\capacitygeneration$. As shown in \cite{brown_decreasing_2020} and in detail in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition}.   
\end{subequations}

@@

Let $L \subset I$ be the set of transmission lines in the system, these may include Alternating Current (AC) as well as Directed Current (DC) lines. Further let $\flow = s_{\ell,t}$ represent the power flow on line $\ell \in L$.  If the OPEX of the transmission system is taken into account in $\totalcost$ (these are often neglected in power system models), these may be approximated by $\opexflow_\ell = \sum_t o_\ell |\flow|$, that is, a cost weighted sum of the net flow on line $\ell$. Again this stands in contrast to the CAPEX which not a function of $\flow$ but of the transmission capacity $\capacityflow$. It limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
At the cost-optimum, the two constraints yield the shadow prices $\muupperflow$ and $\mulowerflow$. We use the relation that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow (for details see \cref{sec:zero_profit_flow}) 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, the shadow prices $\muupperflow$ and $\mulowerflow$ can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, a non-zero values indicate that \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint} are bound and therefore that the congestion on line $\ell$ at time $t$ is imposing costs to the system.

@@

Let $R \in S$ denote all storages in the system. In a simplified storage model, $\capacitystorage$ limits the storage dispatch $\storagedispatch$ and charging $\storagecharge$. Further it limits the maximal storage capacity $\storagesoc$ by a fix ratio $h_r$, denoting the maximum hours at full discharge. The storage $r$ dispatches power with efficiency $\efficiencydispatch$, charges power with efficiency $\efficiencycharge$ and preserves power from one time step $t$ to the next, $t+1$, with an efficiency of $\efficiencysoc$. In \cref{sec:zero_profit_storage_units} we formulate the mathematical details. The OPEX term which adds to the objective function is given by 
\begin{align}
    \opexstorage = \sum_r o_r \storagedispatch 
\end{align}
Using the result of \cite{brown_decreasing_2020} the CAPEX can be related to the operation of a storage unit $r$ through
\begin{align}
    \notag
    \capexstorage =& \capitalpricestorage \, \capacitystorage \\
    \notag
    =& \sum_t \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \storagedispatch \\
    &- \sum_t \lmp \incidencestorage  \storagecharge \Forall{r} 
    \label{eq:no_profit_capex_storage}
\end{align}
where $\muupperstoragedispatch$ and $\mulowerstoragedispatch$ are the shadow prices of the upper and lower dispatch capacity bound and $\mustateofcharge$ is the shadow price of the energy balance constraint. Following the considerations in \cref{sec:zero_profit_storage_units} we restrict to the revenue from dispatched power, \ie the first term in \cref{eq:no_profit_capex_storage}, for the cost allocation.

@@

Dispatch and flow in a system can be considered as a superposition of individual contributions of nodes and assets. In order to artificially quantify these contribution, the literature provides various methods, named flow allocation schemes. Each of these follow a specific set of assumptions which result in peer-to-peer allocations $A_{m \rightarrow n}$. That is a measure for the power supplied at node $m$ and consumed at node $n$.

In the following, we resort to one specific flow allocation scheme Average Participation (AP), also known as Flow Tracing. The basic idea is to trace the power injection at bus $m$ through the network while following the real power flow on transmission lines and applying the principal of proportional sharing. At each bus, including the starting bus $m$, the traced flow might be mixed with incoming power flows from other buses. As soon as the power is absorbed or flowing out of a bus, the traced flow originating from $m$ splits in the same proportion as the total flow. This assumption leads to regionally confined allocation $A_{m \rightarrow n}$ based on a straightforward principal. A mathematical formulation of the AP scheme is documented in \cref{sec:net_ap}. For a detailed comparison with other scheme we refer to [].

The peer-to-peer allocations $\allocatepeer$ fulfill some basic properties. On the one hand it allocates the all power productions at time $t$, \ie when summing over all receiving nodes $n$ the allocations yield the gross power generation of producing assets $i \in S \cup R$ (generators and storages) attached to $m$, denoted by $\nodalgeneration[m]$. Mathematically this translates to    
\begin{align}         
    \nodalgeneration[m] = 
    \sum_{i \in S \cup R} \incidenceasset[m] \, \state = \sum_n \allocatepeer
    \label{eq:nodalgeneration}
\end{align}
where $\incidenceasset[m]$ is 1 if asset $i$ is attached to bus $m$ and zero otherwise. On the other hand, when summing over all supplying nodes, the allocation  $\allocatepeer$ yields the gross nodal demand $\demand$ at node $n$ and time $t$, which leaves us with  
\begin{align}
    \demand = \sum_m \allocatepeer 
\end{align}
As the standard formulation implies, we assume that only net power production of $m$ is allocated to other buses. That is, if the nodal generation $\nodalgeneration[m]$ does not exceed the nodal demand $\demand[m]$, all of it is allocated to local consumers, \ie $\allocatepeer[m \rightarrow m] = \min\left(\nodalgeneration[m], \demand[m]\right)$. 
The principal of proportional sharing sets the allocation of a single producing asset $i \in S \cup R$ proportional to the nodal allocation $\allocatepeer$ and weights it by the share $w_{i,t} = \incidenceasset[m] \state / \nodalgeneration[m]$ that asset $i$ contributes to the nodal generation $\nodalgeneration[m]$, leading to 
\begin{align}
    \allocatestate = w_{i,t} \allocatepeer \Forall{i \in S \cup R, n , t}
    \label{eq:allocate_production}
\end{align}

This relation builds one crucial key to the cost allocation. It states how much of the power produced by asset $i \in S \cup R$ is finally consumed by consumers at $n$.

Yet, we didn't touch the allocation of transporting assets $i \in L$ to consumers. Note that the traced flow based in the AP scheme does obey the Kirchhoff Current Law but not the Kirchhoff Voltage Law, as already pointed out in \cite{galiana_transmission_2003}. However, as we show later a consideration of both laws is needed in order to align the allocated costs with the optimized Locational Marginal Prices. To tackle this, let $\ptdf$ denote a linear mapping between the injection $\left(\nodalgeneration[m] - \demand[m]\right)$ and the flow $\flow$, such that  
\begin{align}
 \flow  = \sum_m \ptdf[m]\, \left(\nodalgeneration[m] - \demand[m]\right) \Forall{\ell \in L, t}
 \label{eq:power_flow_equation}  
\end{align}
Usually, $\ptdf$ is given by the Power Transfer Distribution Factors (PTDF) which indicate the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $m$. For transport models or networks with High Voltage Directed Current (HVDC) lines, these can retrospectively be calculated or extended using the formulation presented in \cite{hofmann_flow_2020}.
Inserting \cref{eq:nodalgeneration} into \cref{eq:power_flow_equation} and expanding the sum yields 
\begin{align}
    \allocateflow = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n m} \demand \right) \Forall{\ell \in L, n, t}
 \label{eq:allocate_flow}
\end{align}
Complementary to \cref{eq:allocate_production}, this allocation indicates the power flow on line $\ell$ and time $t$ which is finally consumed by $\demand$.

@@

Using the presented relations, we are able to straightforwardly define the full cost allocation. Therefore, the expanded sum in \cref{eq:decompose_state} is inserted in \cref{eq:cost_decomposition} which leads to 
\begin{align}
    \allocatecost = \costfactor \, \allocatestate
    \label{eq:cost_allocation}
\end{align}
By default, all cost of a cost term $\cost_i$ associated with asset $i$ are allocated, thus $\cost_i = \sum_n \allocatecost$; likewise all system costs are distributed, $\totalcost = \sum_{\circ, i, n, t} \allocatecost$.

The cost allocation entails a further important property. In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the change of costs for an incremental increase of electricity demand $\demand$ at node $n$ and time $t$ \cite{schweppe_spot_1988}. Mathematically this translates to the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$, $\lmp = \partial \totalcost /\partial \demand$. At the optimum this quantity is given by a shadow price of the nodal balance constraint (see \cref{sec:lmp} for details). Now, when summing over all assets $i$ and cost terms $\circ$, the cost allocation yields 
\begin{align}
    \payment = \sum_{\circ, i} \allocatecost  = \lmp\, \demand  \Forall{n,t}
    \label{eq:payment}
\end{align} 
which we refer to as the nodal payment or payment. It exactly matches the payment determined by the LMP $\lmp$. Hence, \Cref{eq:payment} which is in detailed proved in \cref{sec:proof_equivalence} shows that the cost allocation is embedded in the nodal pricing scheme implied by the LMP.

@@

\begin{subequations}
In real-world setups, generators, lines or other assets can often only be built up to a certain limit. This might be due to land use restrictions or social acceptance considerations. 
When constraining the capacity $\capacity$ to an upper limit $\capacityupper$, in the form of 
\begin{align}
    \capacity - \capacityupper \le 0 \resultsin{\muuppernom} \Forall{i \in I}
\label{eq:capacityexpansionmaximum},
\end{align}
the zero profit condition alters as soon as the constraint becomes binding. Then, asset $i$ is payed an additional scarcity rent 
\begin{align}
    \scarcitycost_i = - \muuppernom \capacity \Forall{i \in I}
    \label{eq:scarcitycost}
\end{align}
This rent may account for different possible realms, as for example the increased market price due in higher competed areas or additional costs for social or environmental compensation. To end this, the share in $\allocatecost$ which consumers pay for the scarcity rent can be recalculated by a correct weighting the shadow price $\muuppergenerationnom$ with the capital price $c_i$, leading to 
\begin{align}
    \allocatescarcitycost = \dfrac{\muuppernom}{c_i + \muuppernom} \, \allocatecost \Forall{i}
\end{align}
\end{subequations}

@@

In the presented work, we restricted the application to long term investment models with perfect foresight. However, the cost allocation can as well be applied to short term planning models with fixed capacities. The revenue from the capacity limits then builds basis for amortization and future investments.

@@

All figures and number of the plot can be reproduced by using the \textit{snakemake} workflow in [].

>>>>

\newcommand{\allocatecost}[1][n \rightarrow i]{\cost_{#1, t}}
\newcommand{\allocatecapex}[1][n \rightarrow s,t]{\mathcal{I}_{#1}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s,t]{\capexgeneration_{#1}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell,t]{\capexflow_{#1}}
\newcommand{\allocatecapexstorage}[1][n \rightarrow r, t]{\capexstorage_{#1}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\opex_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\emissioncost_{#1,t}}
\newcommand{\allocatescarcitycost}[1][n \rightarrow i]{\scarcitycost_{#1,t}}
\newcommand{\allocatesubsidycost}[1][n \rightarrow i]{\subsidycost_{#1,t}}

@@

    \begin{abstract}
        Cost minimization of power systems leads to perfectly scheduled operations and investments while securing all power demands. 
        System needs reports from regional, national and international agencies increasingly rely on cost minimization techniques. However, cost impact to single regions and thefore the question of who profits to what extent from an investment is often neglected. The literature provides multiple cost allocation methods dedicated to this problem, yet they hardly suit for large networks and an integration of all costs. 
        This work introduces a peer-to-peer allocation of all system costs in optimized models. Combining basic relations from economic optimization with flow allocation, its cost assignments are locally constraint while still being aligned with the Locational Marginal Prices at the optimum. It is showcased by means of an optimized near-future scenario of the German power system. 
        \vtop{%
         \vskip0pt
          \hbox{%
          \includegraphics[width=\linewidth]{graphical_abstract.png}
          }%
        }

@@

\subsection*{Highlights}
\begin{itemize}
    \item In long-term equilibria a network asset recovers its variable and investment cost from its operation based revenue
    \item Flow tracing is used as a basis to allocate the operation of assets to consumers. 
    \item Allocated flows have to be reshaped such that the Kirchhoff Current Law and the Kirchhoff Voltage Law are respected. 
    \item Using operational and shadow prices from constraints, all costs are assigned in an P2P manner.
    \item The cost assignments are locally constraint and in alignment with the nodal pricing scheme of the optimum.    
    % \item Upper and lower capacity expansion limits modify the revenue per asset, which however correctly reflects in the cost allocation
    \item In a low-carbon scenario for Germany, regions with high renewable potentials profit from investments compensated by remote buses.   
\end{itemize}

@@

Today's power systems are subject to a deep and ongoing transformation. The shift from controllable to variable, weather-driven power generation as well as the constant improvement and innovation of technology require rigorous system planning and international cooperation \cite{pfenninger_energy_2014,schlachtberger_benefits_2017}. The core of the challenge manifests in the system costs. Firstly, these should be as small as possible while meeting ecological and techno-economic standards. Secondly, they must be distributed in a fair and transparent manner among all agents in the power system. In this respect, it is central to identify the drivers of costs and address them appropriately.  
To this end, power system models are a valuable and widely used tool \cite{pfenninger_energy_2014,bazmi_sustainable_2011,pereira_generation_2017,brown_sectoral_2019}. However, the question of how and, more importantly, upon what grounds costs are distributed is often left open. \\
This study bridges this gap. On the basis of an optimized network the operational state of each time step and corresponding characteristics are in detailed considered to allocate all system costs. The approach builds upon two concepts: Firstly, according to the zero-profit condition at the cost-optimum, the revenue of an network asset exactly matches its operational and capital expenditures. Secondly, flow tracing, following Bialek's Average Participation (AP) \cite{bialek_tracing_1996}, is used to assign the operation of single network assets to consumers in an plausible and locally constraint manner. 
Combining these two relations allows for an transparent allocation of all operational (OPEX) and capital expenditures (CAPEX) to the consumers in the network. \\
Flow-based cost allocation was already discussed and performed in various works \cite{galiana_transmission_2003,shahidehpour_market_2002,meng_investigation_2007,tranberg_flow-based_2018,nikoukar_transmission_2012,arabali_pricing_2012,wu_locational_2005}. Shahidehpour et al. give a deep insight into allocating congestion cost and transmission investments to market participants \cite{shahidehpour_market_2002} with different allocation techniques. Based on Generation Shift Factors, \ie the marginal contribution of generators to a flow on line, the Locational Marginal Price (LMP) can be represented as a superposition of LMP at the reference bus, the price for congestion and a price for losses. The approach in \cite{meng_investigation_2007} expands this relation for contributions based on the AP scheme, which allows for an accurate estimation of the LMP, however this does not reflect the exact LMP of an optimized network. A similar approach was taken in \cite{tranberg_flow-based_2018} where electricity prices of a non-optimal power dispatch were allocated using flow tracing.

@@

The first section which presents the full method splits into the presentation of: the zero-profit condition for different kind of assets (\cref{sec:zero_profit_rule});  the AP allocation scheme and derived allocations from asset to consumer (\cref{sec:dispatch_allocations}); the cost allocation and the impact of additional constraints (\cref{sec:general_scheme,sec:design_constraints}); a numerical example (\cref{sec:numerical_example}). The second part in \cref{sec:application_case} applies the cost allocation to a optimized German power system with a high share of renewable power generation and evaluates the allocated costs.

@@

\subsubsection*{Generators}
\begin{subequations}
Let $S \subseteq I$ be the set of generators in the network, such that $\generation = s_{s,t}$ describes the power production of generator $s \in S$. The OPEX occasioned by generator $s$ is given by a cost-weighted sum of the production, thus 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
In case a fix price for emissions $\emissionprice$ in \euro\, per tonne-CO$_2$ equivalents, is assumed, a further the cost term per generator $s$,   
\begin{align}
 \emissioncost_s = \emissionprice \, \sum_t  \emission \, \generation
\end{align}
adds to $\totalcost$. Here, $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour\, of generator $s$. In contrast to OPEX and emission costs, the CAPEX of $S$ are not a function of the production $\generation$, but of the actual installed capacity $\capacitygeneration$ of generator $s$. In the optimization it limits the generation $\generation$ in the form of 
\begin{align}
\generation - \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}
\end{align}
The constraint yields a shadow-price of $\muuppergeneration$ in literature often denoted as the Quality of Supply \cite{schweppe_spot_1988}. It can be interpreted as the price per MW that \cref{eq:upper_generation_capacity_constraint} imposes to the system. If $\muuppergeneration$  is bigger than zero, the constraint is binding, which pushes investments in $\capacitygeneration$. As shown in \cite{brown_decreasing_2020} and in detail in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition}.   
\end{subequations}

@@

Let $L \subset I$ be the set of transmission lines in the system, these may include Alternating Current (AC) as well as Directed Current (DC) lines. Further let $\flow = s_{\ell,t}$ represent the power flow on line $\ell \in L$.  If the OPEX of the transmission system is taken into account in $\totalcost$ (these are often neglected in power system models), these may be approximated by $\opexflow_\ell = \sum_t o_\ell |\flow|$, that is, a cost weighted sum of the net flow on line $\ell$. Again this stands in contrast to the CAPEX which not a function of $\flow$ but of the transmission capacity $\capacityflow$. It limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
At the cost-optimum, the two constraints yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the relation that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow (for details see \cref{sec:zero_profit_flow}) 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
The shadow prices $\muupperflow$ and $\mulowerflow$ can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, a non-zero values indicate that \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint} are bound and therefore that the congestion on line $\ell$ at time $t$ is imposing costs to the system.

@@

Let $R \subset I$ denote all storages in the system. In a simplified storage model, $\capacitystorage$ limits the storage dispatch $\storagedispatch$ and charging $\storagecharge$. Further it limits the maximal storage capacity $\storagesoc$ by a fix ratio $h_r$, denoting the maximum hours at full discharge. The storage $r$ dispatches power with efficiency $\efficiencydispatch$, charges power with efficiency $\efficiencycharge$ and preserves power from one time step $t$ to the next, $t+1$, with an efficiency of $\efficiencysoc$. In \cref{sec:zero_profit_storage_units} we formulate the mathematical details. The OPEX which adds to $\totalcost$ is given by 
\begin{align}
    \opexstorage = \sum_r o_r \storagedispatch 
\end{align}
Using the result of \cite{brown_decreasing_2020} the CAPEX can be related to the operation of a storage unit $r$ through
\begin{align}
    \notag
    \capexstorage =& \capitalpricestorage \, \capacitystorage \\
    \notag
    =& \sum_t \left(\muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge \right) \storagedispatch \\
    &- \sum_t \lmp \incidencestorage  \storagecharge \Forall{r} 
    \label{eq:no_profit_capex_storage}
\end{align}
where $\muupperstoragedispatch$ and $\mulowerstoragedispatch$ are the shadow prices of the upper and lower dispatch capacity bound and $\mustateofcharge$ is the shadow price of the energy balance constraint. Following the considerations in \cref{sec:zero_profit_storage_units} we restrict to the revenue from dispatched power, \ie the first term in \cref{eq:no_profit_capex_storage}, for the cost allocation.

@@

Dispatch and flow in a system can be considered as a superposition of individual contributions of nodes and assets. In order to artificially quantify these contribution, the literature provides various methods, named flow allocation schemes. Each of these follow a specific set of assumptions which result in peer-to-peer allocations $A_{m \rightarrow n}$. That is a measure for the power produced at node $m$ and consumed at node $n$.

In the following, we resort to one specific flow allocation scheme Average Participation (AP), also known as Flow Tracing. The basic idea is to trace the power injection at bus $m$ through the network while following the real power flow on transmission lines and applying the principal of proportional sharing. At each bus, including the starting bus $m$, the traced flow might be mixed with incoming power flows from other buses. As soon as the power is absorbed or flowing out of a bus, the traced flow originating from $m$ splits in the same proportion as the total flow. This assumption leads to regionally confined allocation $A_{m \rightarrow n}$ based on a straightforward principal. A mathematical formulation of the AP scheme is documented in \cref{sec:net_ap}. For a detailed comparison with other scheme we refer to \cite{hofmann_techno-economic_2020}.

The peer-to-peer allocations $\allocatepeer$ fulfill some basic properties. On the one hand it allocates the all power productions at time $t$, \ie when summing over all receiving nodes $n$ the allocations yield the gross power generation $\nodalgeneration[m]$ of producing assets $i \in S \cup R$ (generators and storages) attached to $m$. Mathematically this translates to    
\begin{align}         
    \nodalgeneration[m] = 
    \sum_{i \in S \cup R} \incidenceasset[m] \, \state = \sum_n \allocatepeer
    \label{eq:nodalgeneration}
\end{align}
where $\incidenceasset[m]$ is 1 if asset $i$ is attached to bus $m$ and zero otherwise. On the other hand, when summing over all supplying nodes, the allocation  $\allocatepeer$ yields the gross nodal demand $\demand$ at node $n$ and time $t$, which leaves us with  
\begin{align}
    \demand = \sum_m \allocatepeer 
\end{align}
As the standard formulation implies, we assume that only net power production of $m$ is allocated to other buses, \ie $\allocatepeer[m \rightarrow m] = \min\left(\nodalgeneration[m], \demand[m]\right)$. Like this nodal generations which do not exceed the nodal demand are completely allocated to local consumers. 
The principal of proportional sharing sets the allocation of a single producing asset $i \in S \cup R$ proportional to the nodal allocation $\allocatepeer$ and weights it by the share $w_{i,t} = \incidenceasset[m] \state / \nodalgeneration[m]$ that asset $i$ contributes to the nodal generation $\nodalgeneration[m]$, leading to 
\begin{align}
    \allocatestate = w_{i,t} \allocatepeer \Forall{i \in S \cup R, n , t}
    \label{eq:allocate_production}
\end{align}
% 
This relation states how much of the power produced by asset $i \in S \cup R$ is finally consumed by consumers at $n$.

Yet, we didn't touch the allocation of transporting assets $i \in L$ to consumers. Note that the traced flow based in the AP scheme does obey the Kirchhoff Current Law but not the Kirchhoff Voltage Law, as already pointed out in \cite{galiana_transmission_2003}. However, as we show later a consideration of both laws is needed in order to align the allocated costs with the optimized Locational Marginal Prices. To tackle this, let $\ptdf$ denote a linear mapping between the injection $\left(\nodalgeneration[m] - \demand[m]\right)$ and the flow $\flow$, such that  
\begin{align}
 \flow  = \sum_m \ptdf[m]\, \left(\nodalgeneration[m] - \demand[m]\right) \Forall{\ell \in L, t}
 \label{eq:power_flow_equation}  
\end{align}
Usually, $\ptdf$ is given by the Power Transfer Distribution Factors (PTDF) which indicate the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $m$. For transport models or networks with High Voltage Directed Current (HVDC) lines, these can retrospectively be calculated using the formulation presented in \cite{hofmann_flow_2020}.
Inserting \cref{eq:nodalgeneration} into \cref{eq:power_flow_equation} and expanding the sum yields 
\begin{align}
    \allocateflow = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n m} \demand \right) \Forall{\ell \in L, n, t}
 \label{eq:allocate_flow}
\end{align}
Complementary to \cref{eq:allocate_production}, this allocation indicates the power flow on line $\ell$ and time $t$ which is finally consumed by $\demand$.

@@

Using the presented relations, we are able to straightforwardly define the cost allocation. Therefore, we insert \cref{eq:decompose_state} in \cref{eq:cost_decomposition} and decompose the sum, which leads to 
\begin{align}
    \allocatecost = \costfactor \, \allocatestate
    \label{eq:cost_allocation}
\end{align}
By default, all cost of a cost term $\cost_i$ associated with asset $i$ are allocated, thus $\cost_i = \sum_n \allocatecost$; likewise all system costs are distributed, $\totalcost = \sum_{\circ, i, n, t} \allocatecost$.

The cost allocation entails a further important property. In a cost-optimal setup with minimized $\totalcost$, the LMP describes the change of costs for an incremental increase of electricity demand $\demand$ at node $n$ and time $t$ \cite{schweppe_spot_1988}. Mathematically this translates to the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$, $\lmp = \partial \totalcost /\partial \demand$. At the optimum this quantity is given by a shadow price of the nodal balance constraint (see \cref{sec:lmp} for details). Now, when summing over all assets $i$ and cost terms $\circ$, the cost allocation yields 
\begin{align}
    \payment = \sum_{\circ, i} \allocatecost  = \lmp\, \demand  \Forall{n,t}
    \label{eq:payment}
\end{align} 
which we refer to as the nodal payment or payment.  
This relation which is in detailed proven in \cref{sec:proof_equivalence} shows that the cost allocation is embedded in the optimized nodal pricing scheme, \ie the nodal payment exactly match the payment determined by the LMP $\lmp$.

@@

\begin{subequations}
In real-world setups, generators, lines or other assets can often only be built up to a certain limit. This might be due to land use restrictions or social acceptance considerations. 
When constraining the capacity $\capacity$ to an upper limit $\capacityupper$, in the form of 
\begin{align}
    \capacity - \capacityupper \le 0 \resultsin{\muuppernom} \Forall{i \in I}
\label{eq:capacityexpansionmaximum},
\end{align}
the zero profit condition alters as soon as the constraint becomes binding. Then, asset $i$ is payed an additional scarcity rent 
\begin{align}
    \scarcitycost_i = - \muuppernom \capacity \Forall{i \in I}
    \label{eq:scarcitycost}
\end{align}
This rent may account for different possible realms, as for example the increased market price in higher competed areas or additional costs for social or environmental compensation. To end this, the share in $\allocatecost$ which consumers pay for the scarcity rent can be recalculated by a correct weighting of the shadow price $\muuppergenerationnom$ with the capital price $c_i$, leading to 
\begin{align}
    \allocatescarcitycost = \dfrac{\muuppernom}{c_i + \muuppernom} \, \allocatecost \Forall{i}
\end{align}
\end{subequations}

@@

We restricted the application to long term investment models with perfect foresight. However, the cost allocation can as well be applied to short term planning models with fixed capacities. The revenue from the capacity limits then builds basis for amortization and future investments.

@@

All figures and data points can be reproduced by using the \textit{snakemake} workflow in \cite{hofmann_pypsa-costallocation_2020}.
